Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-03-25 jss/EpiCompare-paper/EpiCompare-paper.Rmd
updates to section 2 by Ben (after shannon's updates)

title:
  formatted: "Time invariant analysis of epidemics with \\pkg{EpiCompare}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Time invariant analysis of epidemics with EpiCompare"
  # For running headers, if needed
  short:     "\\pkg{EpiCompare}"
abstract: >
  We present \pkg{EpiCompare}, an \proglang{R} package that suppliments and enhances current infectious disease analysis pipelines and encourages comparisons across models and epidemics. A major contribution of this work is the set of novel \textit{time-invariate} tools for model and epidemic comparisons - including time-invariate prediction bands. \pkg{EpiCompare} embraces \proglang{R}'s \textit{tidy} coding style to make adoption of the package easier and analysis faster. This paper provides an overview of both the tools in and intuition behind \pkg{EpiCompare} and a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
  \usepackage{amsthm}
  \usepackage{afterpage}
  \usepackage[normalem]{ulem} 
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
classoption: shortnames
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: EpiCompare.bib  #Also not included in template by default. 
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{xcolor}
  - \usepackage{rotating}
---

@@

## Motivating time-invariance through $R_0$ {short-title="motivating through R0" #r0}

Time-invariance analysis, as it appears in \pkg{EpiCompare}, \sout{can be seen as an attempt to escape} \shannon{bypasses}  many difficulties \shannon{in} comparing different epidemics. With time-invariant analysis, comparing the decades-long outbreak of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is \sout{still} possible. Time-dependent problems can arise when estimation of certain epidemiological parameters, including the reproduction number $R_0$.

\ben{
Story: 
(1) basic definition of $R_0$, (2) emphasis that $R_0$ captures a lot of information about an epidemic. Given 3 number example and discuss how one would usually interpret what the differences in $R_0$ meant (3) maybe connect to \# of $R_0$ estimates for COVID-19 (Aronson et al 2020). 
}

$R_0$ is probably the most famous time-invariant numerical summary of an epidemic and is \shannon{often} associated with the Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2001}.  \shannon{$R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a naive population \citep{anderson1992}. $R_0$ can be a difficult quantity to estimate and is typically sensitive to time-based parameters such as the beginning and end of an epidemic \citep{Gallagher2020}. } To demonstrate the difficulty of discerning $R_0$in a time-dependent analysis, we \shannon{first} introduce the \citet{Kermack1927}’s common SIR model. This model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,

@@

## Visualizing $R_0$ {short-title="visualizing r0" #vis-r0}

With regards to traditional epidemic $state$ vs. $time$ plots, when \sout{comparing}\shannon{examining} different epidemics their $R_0$s are often impossible to compare. This can be seen in a basic example, where we have two different epidemics that start in two populations of 1000 people with 10 individuals \shannon{initially} infected, but with \sout{different parameters} \shannon{different infection and recovery rates (see Eq. \ref{eq:sir-ODE}}). Specifically suppose that $\beta_1, \gamma_1 = (a,b)$ and $\beta_2, \gamma_2 = (ca,cb)$ where $a, b, c > 0$. Further suppose we observe both of these epidemics for 15 days. The epidemic trajectories are shown in the $state$ vs. time plots in Figure \ref{fig:different-scales-standard}.  At a glance, we may \sout{assume} \shannon{believe} that Model 1 has a larger $R_0$ than Model 2 because the peak of infection occurs more quickly than in Model 2.  On the other hand, we may \sout{think} \shannon{believe} Model 2 has a larger $R_0$ because we may think the number of infections in that model has not yet peaked at time 15. Yet, both epidemics have the same $R_0$ of 2.

@@

However, time-invariant centric analysis allows us to visualize $R_0$ for the Kermack and McKendrick SIR model.  We call the epidemic trajectory of number of individuals in each state a \textit{filament}.  Mathematically, a filament $\psi$ for the SIR model  $\psi$ this is the set of points of the number in each state for each time point,
$$
\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t=0:T}.
$$

@@

We visualize the two epidemics in a global, ternary view in Figure \ref{fig:different-scales-tern}.  Without getting into too much detail of the intricacies of this plot, we immediately see the points of the two filaments $\psi$ seem to form the same trajectory.  Now, it is much clearer that Model 2 is following the same trajectory as Model 1 but is not as far along in the infection process.  We can see there is something fundamentally linking these two different epidemics, and this fundamental link turns out to be $R_0$.

We can show this mathematically if we let our two epidemics be presented as $\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}$, $\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}$ respectively.  As with the example, assume both models have the same initial values $(S(0), I(0), R(0))$, and let $R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$. And define $a>0$ to be the relative scalar such that $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$.

@@

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.  The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model is also a point on the epidemic path of the second SIR model.

@@

##  Beyond $R_0$ and Kermack’s and McKendrick SIR Models {short-title="Beyond R0 and SIR" #beyond-r0-sir}

Through $R_0$ \shannon{example}, we see that treating epidemics more like a \sout{function of the time points in some higher dimensional space (nicely visualizable with a ternary plot for the SIR models)} \shannon{filaments embedded in a lower dimensional space}, then we can better compare the overall structure of the epidemic and \sout{how it impacted the population} \shannon{see how the population was directly impacted}. \sout{Without knowing the generative structure underlying these two epidemics}, \shannon{When the underlying generative model is unknown for epidemic trajectories}, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states.

@@

\afterpage{\clearpage}
\begin{sidewaysfigure}[!ht]
    \centering
    \includegraphics[width = 1\textwidth]{images/pipeline2.pdf}
    \caption{How \pkg{EpiCompare} supplements and aids in the epidemiological data analysis pipeline.}
    \label{fig:pipeline2}
\end{sidewaysfigure}

>>>>

title:
  formatted: "Time invariant analysis of epidemics with \\pkg{EpiCompare}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Time invariant analysis of epidemics with EpiCompare"
  # For running headers, if needed
  short:     "\\pkg{EpiCompare}"
abstract: >
  We present \pkg{EpiCompare}, an \proglang{R} package that suppliments and enhances current infectious disease analysis pipelines and encourages comparisons across models and epidemics. A major contribution of this work is the set of novel \textit{time-invariate} tools for model and epidemic comparisons - including time-invariate prediction bands. \pkg{EpiCompare} embraces \proglang{R}'s \textit{tidy} coding style to make adoption of the package easier and analysis faster. This paper provides an overview of both the tools in and intuition behind \pkg{EpiCompare} and a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{amsthm}
  \usepackage{afterpage}
  \usepackage[normalem]{ulem} 
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
classoption: shortnames
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: EpiCompare.bib  #Also not included in template by default. 
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{xcolor}
  - \usepackage{rotating}
---

@@

## Motivating time-invariance through $R_0$^[\ben{Is this still a relevant title for this section?}] {short-title="motivating through R0" #r0}

Time-invariance analysis, as it appears in \pkg{EpiCompare}, \sout{can be seen as an attempt to escape} \shannon{bypasses}  many difficulties \shannon{in} comparing different epidemics. With time-invariant analysis, comparing the decades-long outbreak of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is \sout{still} possible. Time-dependent problems can arise when estimation of certain epidemiological parameters, including the reproduction number $R_0$. \ben{We will use $R_0$ to motivate the usefulness of time-invariaten analysis in this section.}

<!-- \ben{ -->
<!-- Story:  -->
<!-- (1) basic definition of $R_0$, (2) emphasis that $R_0$ captures a lot of information about an epidemic. Given 3 number example and discuss how one would usually interpret what the differences in $R_0$ meant (3) maybe connect to \# of $R_0$ estimates for COVID-19 (Aronson et al 2020).  -->
<!-- } -->

$R_0$ is probably the most famous time-invariant numerical summary of an epidemic and is \shannon{often} associated with the Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2001}.  \shannon{$R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a \ben{\sout{naive} completely susceptible}\footnote{\ben{[Ben says: I feel like naive is unclear - even if that's the technical term.]}} population \citep{anderson1992}. $R_0$ can be a difficult quantity to estimate and is typically sensitive to time-based parameters such as the beginning and end of an epidemic \citep{Gallagher2020}. }\ben{[Ben says: could we expand on this difficulty / show it really is difficult with 1 more sentence?]} To demonstrate the difficulty of discerning $R_0$in a time-dependent analysis, we \shannon{first} introduce the \citet{Kermack1927}’s \ben{\sout{common}}\footnote{\ben{[Ben says: is it common? maybe just remove it?]}} SIR model. This model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,

@@

## \ben{Time invariant visualization of differences in $R_0}\sout{ Visualizing $R_0$} {short-title="visualizing r0" #vis-r0}

\ben{[Ben says: It's unclear to me why we have a subtitle here - isn't it just more motivation of time-invariant anlysis with $R_0$? Also, I feel like the story is weak here. The point is to leverage $R_0$ to show the value of time-invariant analysis - this seems a bit more like just discussing properties of $R_0$. In the follow rewrite I use "[" and "]" to indicate that this is a section from your earlier draft.]}

\ben{[Ben says: replacement of following paragraph: presentation rearranged to emphasis the figure first (when presenting the problem), and then discussing it.]} Traditional $state$ vs. $time$ plots, like seen in Fig. \ref{fig:different-scales-standard}, present information about epidemics in a time-dependent and piece-wise fashion, and make comparing different epidemic $R_0$s impossible. In Fig. \ref{fig:different-scales-standard} we compare [two different epidemics that start in two populations of 1000 people with 10 individuals initially infected, but with different infection and recovery rates (see Eq. \ref{eq:sir-ode})]. More specifically, we've observed both epidemics for 15 days, with the first epidemic\footnote{In the plot it says model - probably want to say epidemic?} (in red) having parameters $\beta_1, \gamma_1 = (1,2)$ and the second epidemic (in blue) having  parameters $\beta_2, \gamma_2 = (2,4)$. [At a glance, we may believe that \sout{Model}\ben{epidemic} 1 has a larger $R_0$ than \sout{Model}\ben{epidemic} 2 because the peak of infection occurs more quickly than in Model 2 \ben{(remember $R_0$ relates to the rate of infection)}.  On the other hand, we may believe \sout{Model}\ben{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \sout{model}\ben{epidemic} has not yet peaked at time 15. Yet, both epidemics have the same $R_0$ of 2].

\ben{[Ben says: bring Fig. \ref{fig:different-scales-standard} here.]}

\ben{[Ben says: this paragraph is replaced with the one above]}With regards to traditional epidemic $state$ vs. $time$ plots, when \sout{comparing}\shannon{examining} different epidemics their $R_0$s are often impossible to compare. This can be seen in a basic example, where we have two different epidemics that start in two populations of 1000 people with 10 individuals \shannon{initially} infected, but with \sout{different parameters} \shannon{different infection and recovery rates (see Eq. \ref{eq:sir-ode}}). Specifically suppose that $\beta_1, \gamma_1 = (a,b)$ and $\beta_2, \gamma_2 = (ca,cb)$ where $a, b, c > 0$. Further suppose we observe both of these epidemics for 15 days. The epidemic trajectories are shown in the $state$ vs. $time$ plots in Figure \ref{fig:different-scales-standard}.  At a glance, we may \sout{assume} \shannon{believe} that Model 1 has a larger $R_0$ than Model 2 because the peak of infection occurs more quickly than in Model 2.  On the other hand, we may \sout{think} \shannon{believe} Model 2 has a larger $R_0$ because we may think the number of infections in that model has not yet peaked at time 15. Yet, both epidemics have the same $R_0$ of 2.

@@

\ben{[Ben says: another rewrite of the paragraph below.]} A time-invariant approach to visualizing epidemics, which will be described more below, provides a simple way to compare $R_0$ values of epidemics that are generated from the Kermack and McKendrick SIR model. In these visuals, we treat epidemics as \textit{trajectories}. A trajectory can be mathematically viewed as a set of points is space that have an ordering, and that all points on the line between these order points are also contained in the geometric object. For a SIR epidemic, we can represent the associated filament $\psi$ as

$$
\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]},
$$
where the an mapping $\xi : t \to \mathbb{R}$ that is monotonically increasing would not change the definition of $\psi$, i.e. $\psi_\xi = \psi$ where :

$$
\psi_\xi = \left \{(S(\xi(t)), I(\xi(t)), R(\xi(t))): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]}.
$$
[Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then \ben{all points in} $\psi$ will lay in a \ben{\sout{2d triangular plane, i.e. a ternary plot}} \ben{two-dimensional triangular plane in the three-dimensional space, which can be visualized in a two-dimensional ternary plot}.  As a result, we can visualize the global trajectory of an epidemic in a single \ben{ternary plot \sout{2d plot and ultimately $R_0$}}.  \shannon{[Shannon says: Show pic here?]}]

\ben{[Ben says: this paragraph was replaced with the paragraph above]} However, time-invariant centric analysis allows us to visualize $R_0$ for the Kermack and McKendrick SIR model.  We call the epidemic trajectory of number of individuals in each state a \textit{filament}.  Mathematically, a filament $\psi$ for the SIR model  $\psi$ this is the set of points of the number in each state for each time point,
$$
\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t=0:T}.
$$

@@

\ben{[This section could be a bit less wordy. But generally good.]} We visualize the two epidemics in a global, ternary view in Figure \ref{fig:different-scales-tern}.  Without getting into too much detail of the intricacies of this plot, we immediately see the points of the two filaments $\psi$ seem to form the same trajectory.  Now, it is much clearer that \ben{\sout{Model} epidemic} 2 is following the same trajectory as \ben{\sout{Model} epidemic} 1 but is not as far along in the infection process.  We can see there is something fundamentally linking these two different epidemics, and this fundamental link turns out to be $R_0$.

We can show this mathematically\ben{.\sout{ if we} L}et our two epidemics be presented as $\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}$, $\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}$ respectively.\footnote{\ben{[Ben says: Originally this was a sentence fragment.]}}  As with the example, assume both models have the same initial values $(S(0), I(0), R(0))$, and let $R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$. And define $a>0$ to be the relative scalar such that $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$.

@@

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.  The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model \ben{\sout{is also} can be mapped to} a point on the epidemic path of the second SIR model.\ben{[We've changed how we describe these two objects when they are related to the theorem - is that desirable?]}

@@

##  \ben{Time-invariant analysis} Beyond $R_0$ and Kermack’s and McKendrick SIR Models {short-title="Beyond R0 and SIR" #beyond-r0-sir}

Through \ben{the} $R_0$ \shannon{example}, we see that treating epidemics \ben{\sout{more}} like a \sout{function of the time points in some higher dimensional space (nicely visualizable with a ternary plot for the SIR models)} \shannon{filaments embedded in a lower dimensional space}\ben{\sout{, then we can} allows us to} better compare the overall structure of the epidemic and \sout{how it impacted the population} \shannon{see how the population was directly impacted}. \sout{Without knowing the generative structure underlying these two epidemics}, \shannon{When the underlying generative model is unknown for epidemic trajectories}, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states.

@@

\afterpage{\clearpage}
\begin{sidewaysfigure}[!ht]
    \centering
    \includegraphics[width = 1\textwidth]{images/pipeline2_1.pdf}
    \caption{How \pkg{EpiCompare} supplements and aids in the epidemiological data analysis pipeline.}
    \label{fig:pipeline2}
\end{sidewaysfigure}
